---
title: "About the Data"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(survey)
library(srvyr)
library(here)
library(gt)
library (car)
```

## Data Source

The data used in this analysis was obtained from the Health Information National 
Trends Survey, Cycle 7 (HINTS 7), which was administered by the National Cancer
Institute (NCI). HINTS is a nationally representative, cross-sectional survey of 
U.S. adults aged 18 and older. The survey was conducted between March and 
September 2024 using a mail and web-based design.This is a publicly available 
dataset that can be accessed at [hints.cancer.gov](https://hints.cancer.gov). 


## Loading the data
```{r}
#| label: load-data
#| results: hide

load(here("data", "hints7_public.rda"))
df_raw <- public

nrow(df_raw)
ncol(df_raw)
```
## Variables Used 

The following variables were selected for this analysis.
```{r}
#| label: variable-table

tibble(
  Variable = c(
    "SeekCancerInfo",
    "MisleadingHealthInfo",
    "HealthRecsConflict",
    "CancerTrustDoctor",
    "PreventNotPossible",
    "RaceEthn5",
    "EducA",
    "IncomeRanges_IMP",
    "AgeGrpB",
    "BirthSex",
    "EverHadCancer",
    "HealthInsurance2"
  ),
  Description = c(
    "Whether the respondent sought cancer information in the past year",
    "How much misleading health information seen on social media",
    "How often health recommendations seem to conflict with each other",
    "Level of trust in doctors as a source of cancer information",
    "Belief that cancer prevention is not possible (fatalism)",
    "Race and ethnicity (5 categories)",
    "Highest level of education completed",
    "Annual household income (imputed for missing values)",
    "Age group",
    "Birth sex",
    "Whether the respondent has ever been diagnosed with cancer",
    "Whether the respondent has health insurance"
  ),
  Role = c(
    "Outcome",
    "Key predictor",
    "Key predictor",
    "Key predictor",
    "Key predictor",
    "Demographic",
    "Demographic",
    "Demographic",
    "Demographic",
    "Demographic",
    "Covariate",
    "Covariate"
  ),
  Type = c(
    "Binary (Yes/No)",
    "Ordinal — 4 levels + no social media",
    "Ordinal — Never / Rarely / Often / Very Often",
    "Ordinal — Not at all / A little / Some / A lot",
    "Ordinal — Strongly agree to Strongly disagree",
    "Categorical",
    "Ordinal — 4 levels",
    "Ordinal — collapsed to 4 levels",
    "Ordinal — 5 groups",
    "Binary (Female/Male)",
    "Binary (Yes/No)",
    "Binary (Yes/No)"
  )
) |>
  gt() |>
  tab_header(title = "Variables used in the analysis") |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#E6F1FB"),
    locations = cells_body(rows = Role == "Outcome")
  ) |>
  tab_style(
    style = cell_fill(color = "#E1F5EE"),
    locations = cells_body(rows = Role == "Key predictor")
  ) |>
  cols_width(
    Description ~ px(280),
    Variable ~ px(160),
    Role ~ px(110),
    Type ~ px(200)
  )
```


## Data Cleaning


### Outcome variable — Cancer information seeking

The outcome variable `SeekCancerInfo` was originally coded as a character 
variable with values "Yes" and "No". This was recoded into a binary numeric 
variable (`seek_cancer_info`) where Yes = 1 and No = 0 for logistic regression.


### Key predictor variables

#### **Information Environment Variables**

Two variables capturing the information environment were converted into 
ordered factors:

- `MisleadingHealthInfo` : how much misleading health information the 
respondent encountered on social media was recoded into `misleading_info` 
with five ordered levels: None, A little, Some, A lot, and I do not use 
social media. The "no social media" category was kept as a separate level 
rather than treated as missing, because it represents a meaningfully 
different group of adults who are not exposed to social media health 
content at all.

- `HealthRecsConflict`: how often the respondent encounters conflicting 
health recommendations was recoded into `recs_conflict` with four ordered 
levels: Never, Rarely, Often, and Very Often.

#### **Attitudinal Variables**

Two attitudinal variables were converted into ordered factors:

- `CancerTrustDoctor`: trust in doctors for cancer information was 
recoded into `trust_doctor` with four levels ordered from least to most 
trust: Not at all, A little, Some, A lot.

- `PreventNotPossible`: agreement with the statement that cancer 
prevention is not possible was recoded into `fatalism` with four levels 
ordered from most to least fatalistic: Strongly agree, Somewhat agree, Somewhat
disagree, Strongly disagree.


### Demographic Variables

Several demographic variables required recoding:

- **Race/ethnicity** (`RaceEthn5`) was converted to a factor and releveled 
so that Non-Hispanic White is the reference category in the regression 
model. 

- **Education** (`EducA`) was converted to an ordered factor with four 
levels: Less than High School, High School Graduate, Some College, and 
College Graduate or More. Less than High School serves as the reference 
category in the regression model.

- **Income** required special handling. The original imputed income variable 
(`IncomeRanges_IMP`) had nine detailed income categories. To improve 
interpretability, these were collapsed into four broader categories: 
Less than \$35,000, \$35,000 to \$74,999, \$75,000 to \$99,999, and \$100,000 
or more. The imputed version of the income variable was used rather than the 
original because it has only 0.2% missing data compared to 12.4% for the 
unimputed version. The NCI used a weighted sequential hot deck imputation 
method to fill in missing income values.

- **Age group** (`AgeGrpB`) was converted to an ordered factor with five 
levels: 18–34, 35–49, 50–64, 65–74, and 75+. The 18–34 age group serves 
as the reference category.

- **Sex** (`BirthSex`) was recoded to retain only Female and Male responses. 
The small number of respondents (n = 44) who responded "Don't know" were 
set to missing because this group was too small for reliable estimation. 
Female is the reference category in the regression model.

- **Cancer history** (`EverHadCancer`) and **health insurance** 
(`HealthInsurance2`) were each recoded as binary factors (No/Yes), with 
No as the reference category in both cases.


### Frustration Variables

Four variables measuring the experience of seeking cancer information 
were recoded using the `recode_hints()` helper function to set invalid 
responses to missing. These variables: `frustration_effort`, 
`frustration_quality`, `frustration_hard`, and `frustration_upset` — 
were only asked of respondents who reported seeking cancer information 
(approximately 50% of the sample), so they are used only in the 
follow-up descriptive analysis restricted to information seekers.

```{r}
#| label: clean-and-design

# Helper function
recode_hints <- function(x) {
  x <- as.character(x)
  ifelse(grepl("Missing|error|Error|Unreadable|Inapplicable", x), NA, x)
}

# Clean data
df <- df_raw |>
  mutate(
    seek_cancer_info = case_when(
      recode_hints(SeekCancerInfo) == "Yes" ~ 1L,
      recode_hints(SeekCancerInfo) == "No"  ~ 0L
    ),
    misleading_info = factor(
      recode_hints(MisleadingHealthInfo),
      levels = c("None", "A little", "Some", "A lot",
                 "I do not use social media")
    ),
    recs_conflict = factor(
      recode_hints(HealthRecsConflict),
      levels = c("Never", "Rarely", "Often", "Very Often")
    ),
    trust_doctor = factor(
      recode_hints(CancerTrustDoctor),
      levels = c("Not at all", "A little", "Some", "A lot")
    ),
    fatalism = factor(
      recode_hints(PreventNotPossible),
      levels = c("Strongly agree", "Somewhat agree",
                 "Somewhat disagree", "Strongly disagree")
    ),
    race = factor(recode_hints(RaceEthn5)),
    race = relevel(race, ref = "Non-Hispanic White"),
    education = factor(
      recode_hints(EducA),
      levels = c("Less than High School", "High School Graduate",
                 "Some College", "College Graduate or More")
    ),
    income = case_when(
      recode_hints(IncomeRanges_IMP) %in%
        c("$0 to $9,999", "$10,000 to $14,999", "$15,000 to $19,999",
          "$20,000 to $34,999") ~ "Less than $35,000",
      recode_hints(IncomeRanges_IMP) %in%
        c("$35,000 to $49,999", "$50,000 to $74,999") ~ "$35,000 to $74,999",
      recode_hints(IncomeRanges_IMP) == "$75,000 to $99,999" ~ "$75,000 to $99,999",
      recode_hints(IncomeRanges_IMP) %in%
        c("$100,000 to $199,999", "$200,000 or more") ~ "$100,000 or more"
    ),
    income = factor(income,
      levels = c("Less than $35,000", "$35,000 to $74,999",
                 "$75,000 to $99,999", "$100,000 or more")),
    age_group = factor(
      recode_hints(AgeGrpB),
      levels = c("18-34", "35-49", "50-64", "65-74", "75+")
    ),
    sex = case_when(
      recode_hints(BirthSex) == "Female" ~ "Female",
      recode_hints(BirthSex) == "Male"   ~ "Male"
    ),
    sex = factor(sex, levels = c("Female", "Male")),
    cancer_history = case_when(
      recode_hints(EverHadCancer) == "Yes" ~ "Yes",
      recode_hints(EverHadCancer) == "No"  ~ "No"
    ),
    cancer_history = factor(cancer_history, levels = c("No", "Yes")),
    insurance = case_when(
      recode_hints(HealthInsurance2) == "Yes" ~ "Yes",
      recode_hints(HealthInsurance2) == "No"  ~ "No"
    ),
    insurance = factor(insurance, levels = c("No", "Yes")),
    frustration_effort  = recode_hints(CancerLotOfEffort),
    frustration_quality = recode_hints(CancerConcernedQuality),
    frustration_hard    = recode_hints(CancerTooHardUnderstand),
    frustration_upset   = recode_hints(CancerFrustrated)
  )

# Survey design
svy <- df |>
  as_survey_rep(
    weights    = PERSON_FINWT0,
    repweights = num_range("PERSON_FINWT", 1:50),
    type       = "JKn",
    scale      = 0.98,
    rscales    = rep(1, 50),
    mse        = TRUE
  )
```


### Handling Missing Values
The negative values in the HINTS dataset represent different types of missing 
data such as "Not Ascertained", "Multiple responses selected in error", " Not 
Applicable", "Skipped due to survey logic". All the negative  values were
recorded to `NA` before the analysis. The HINTS 7 survey uses a complex sampling
design, so to ensure that the estimates of our findings are representative of 
U.S adults, sampling weights were applied using the final sample weight 
(PERSON_FINWT0). Standard errors were calculated using 50 jackknife replicate 
weights (PERSON_FINWT1–PERSON_FINWT50), with a degrees of freedom of 49.

#### Missingness summary
Overall, the amount of missing values for our key variables was low. The outcome
variable (`seek_cancer_info`) had less than 1% missing data. The race/ethnicity 
variable had the highest proportion of missing values at 11.5% (n = 840), 
which may affect the estimates for racial subgroups. All the other variables had 
less than 10% missing values. The missing values were excluded from analyses on 
a variable-by-variable basis.

```{r}
#| label: missingness

tibble(
  Variable = c("seek_cancer_info", "misleading_info", "recs_conflict",
               "trust_doctor", "fatalism", "race", "education",
               "income", "age_group", "sex", "cancer_history", "insurance"),
  `N Missing` = c(
    sum(is.na(df$seek_cancer_info)),
    sum(is.na(df$misleading_info)),
    sum(is.na(df$recs_conflict)),
    sum(is.na(df$trust_doctor)),
    sum(is.na(df$fatalism)),
    sum(is.na(df$race)),
    sum(is.na(df$education)),
    sum(is.na(df$income)),
    sum(is.na(df$age_group)),
    sum(is.na(df$sex)),
    sum(is.na(df$cancer_history)),
    sum(is.na(df$insurance))
  )
) |>
  mutate(`% Missing` = round(`N Missing` / nrow(df) * 100, 1)) |>
  gt() |>
  tab_header(title = "Missing data summary for analysis variables") |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_fill(color = "#FCEBEB"),
    locations = cells_body(rows = `% Missing` > 10)
  )
```




# Eploratory Data Analysis
## Distribution of Key Variables
Before running the regression model, we examined the distribution of 
all the key variables to check for sample sizes, whether the outcome was
balanced and potential issues with rare categories.

### Outcome variable
```{r}

#| label: outcome-distributiion

svy |>
  filter(!is.na(seek_cancer_info)) |>
  group_by(seek_cancer_info) |>
  summarise(
    n    = unweighted(n()),
    prop = survey_prop(vartype = "ci") * 100
  ) |>
  mutate(
    label    = ifelse(seek_cancer_info == 1, "Sought Info", "Did Not Seek"),
    prop     = round(prop, 1),
    prop_low = round(prop_low, 1),
    prop_upp = round(prop_upp, 1)
  ) |>
  select(label, n, prop, prop_low, prop_upp) |>
  gt() |>
  tab_header(title = "Weighted prevalence of cancer information seeking") |>
  cols_label(
    label    ~ "Category",
    n        ~ "Unweighted n",
    prop     ~ "Weighted %",
    prop_low ~ "95% CI Lower",
    prop_upp ~ "95% CI Upper"
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#2c7bb6"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(table.font.size = 13)
```

The outcome variable was well balanced. 51.1% of the participants sought cancer
information and 48.9% did not. 

### Categorical Predictors — Cell Size Check
All the key predictor variables had adequate cell sizes. All the cells had more 
that 10 observations.

```{r}
#| label: cell-sizes

library(gt)

check_cells <- function(var, label) {
  df |>
    filter(!is.na(!!sym(var)), !is.na(seek_cancer_info)) |>
    count(!!sym(var), seek_cancer_info) |>
    group_by(!!sym(var)) |>
    mutate(total = sum(n)) |>
    ungroup() |>
    filter(seek_cancer_info == 1) |>
    select(category = !!sym(var), seekers = n, total) |>
    mutate(
      variable    = label,
      pct_seekers = round(100 * seekers / total, 1)
    ) |>
    select(variable, category, total, seekers, pct_seekers)
}

bind_rows(
  check_cells("race",           "Race/Ethnicity"),
  check_cells("education",      "Education"),
  check_cells("income",         "Income"),
  check_cells("age_group",      "Age Group"),
  check_cells("sex",            "Sex"),
  check_cells("cancer_history", "Cancer History"),
  check_cells("insurance",      "Insurance"),
  check_cells("misleading_info","Misleading Info"),
  check_cells("recs_conflict",  "Conflicting Recs"),
  check_cells("trust_doctor",   "Trust in Doctors"),
  check_cells("fatalism",       "Cancer Fatalism")
) |>
  gt() |>
  tab_header(
    title    = "Cell size check for categorical predictors",
    subtitle = "Minimum recommended cell size: 10 events (seekers)"
  ) |>
  cols_label(
    variable    ~ "Variable",
    category    ~ "Category",
    total       ~ "Total n",
    seekers     ~ "Seekers (n)",
    pct_seekers ~ "% Seekers"
  ) |>
  tab_style(
    style     = cell_fill(color = "#FCEBEB"),
    locations = cells_body(rows = seekers < 10)
  ) |>
  tab_style(
    style     = cell_fill(color = "#2c7bb6"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(table.font.size = 12)
```



### Exploratory Plots
The exploratory plots bellow show the distribution of the key categorical
predictors that were included in the survey-weighted logistic regression model.
Our sample was predominantly Non-Hispanic White and well-educated, with majority 
of the respondents being college graduates or having more than college education. 
A big proportion of the survey participants reported exposure to misleading 
health information and conflicting health recommendations. Household income was 
unevenly distributed across the categories, with majority being on the lower end
reporting less than $ 30,000. Most respondents were insured, and the sample 
was dominated by middle-aged and older adults. These plots show that there was
some diversity across the sociodemographic characteristics. However, the 
regression output should be interpreted with caution because some groups were 
over represented for instance Non-Hispanic, highly educated and insured 
respondents. 

```{r}
#| label: eda-plots
#| fig-cap: "Distribution of key predictor variables"

p1 <- df |>
  filter(!is.na(race)) |>
  count(race) |>
  ggplot(aes(y = fct_reorder(race, n), x = n)) +
  geom_col(fill = "#2c7bb6") +
  labs(x = "Count", y = NULL, title = "Race/Ethnicity") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7))

p2 <- df |>
  filter(!is.na(education)) |>
  count(education) |>
  ggplot(aes(y = fct_reorder(education, n), x = n)) +
  geom_col(fill = "#1a9641") +
  labs(x = "Count", y = NULL, title = "Education") +
  theme_minimal(base_size = 11)

p3 <- df |>
  filter(!is.na(misleading_info)) |>
  count(misleading_info) |>
  ggplot(aes(y = fct_reorder(misleading_info, n), x = n)) +
  geom_col(fill = "#d7191c") +
  labs(x = "Count", y = NULL, title = "Misleading Health Info Exposure") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7))

p4 <- df |>
  filter(!is.na(recs_conflict)) |>
  count(recs_conflict) |>
  ggplot(aes(y = fct_reorder(recs_conflict, n), x = n)) +
  geom_col(fill = "#756bb1") +
  labs(x = "Count", y = NULL, title = "Conflicting Recommendations") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7))

p5 <- df |>
  filter(!is.na(income)) |>
  count(income) |>
  ggplot(aes(y = fct_reorder(income, n), x = n)) +
  geom_col(fill = "#BA7517") +
  labs(x = "Count", y = NULL, title = "Household Income") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7))

p6 <- df |>
  filter(!is.na(fatalism)) |>
  count(fatalism) |>
  ggplot(aes(y = fct_reorder(fatalism, n), x = n)) +
  geom_col(fill = "#D85A30") +
  labs(x = "Count", y = NULL, title = "Cancer Fatalism") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7))

p7 <- df |>
  filter(!is.na(age_group)) |>
  count(age_group) |>
  ggplot(aes(y = fct_reorder(age_group, n), x = n)) +
  geom_col(fill = "#2c7bb6") +
  labs(x = "Count", y = NULL, title = "Age Group") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7))

p8 <- df |>
  filter(!is.na(insurance)) |>
  count(insurance) |>
  ggplot(aes(y = fct_reorder(insurance, n), x = n)) +
  geom_col(fill = "#16A085") +
  labs(x = "Count", y = NULL, title = "Health Insurance") +
  theme_minimal(base_size = 11)



gridExtra::grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol = 2,
                        heights = c(1.2, 1, 1, 1))
```


# Regression assumptions checks
```{r}
#| label: build-model
#| include: false

# Build survey design for model
svy_design_model <- svydesign(
  ids     = ~1,
  weights = ~PERSON_FINWT0,
  data    = df
)

# Fit survey-weighted logistic regression
model <- svyglm(
  seek_cancer_info ~ race + education + income + age_group + sex +
    cancer_history + insurance + misleading_info + recs_conflict +
    trust_doctor + fatalism,
  design = svy_design_model,
  family = quasibinomial()
)
```


## Multicollinearity
We used variance inflation factor (VIF) to assess multicollinearity. All 
the VIF values were close to 1 (range 1.03–1.17) indicating no evidence of 
multicollinearity among predictors.

```{r}
vif(model)
```

## Model fit check
We used the McFadden's Pseudo R-squared to check the model fit. The model had a 
pseudo R-squared value of 0.224 which indicates that it is a good model fit.
The demographic and attitudinal predictors that were included in the model 
helped to improve its overall fit compared to the null model in predicting 
cancer information seeking behavior among U.S. adults.


```{r}
#| warning: false
# Pseudo R-squared (McFadden)
null_model <- svyglm(
  seek_cancer_info ~ 1,
  design = svy_design_model,
  family = quasibinomial()
)

1 - (logLik(model) / logLik(null_model))
```

