---
title: "Who Seeks Cancer Information? Predictors in a Nationally Representative U.S. Sample."
author: "Faith Okaalo"
date: today
bibliography: references.bib
csl: apa.csl
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Introduction

Cancer remains one of the leading causes of morbidity and mortality in the 
United States. It is the second leading cause of deaths in the country after 
heart disease [@siegel2024]. Availability and accessibility of accurate and 
timely information is necessary for prevention, early detection and treatment 
of cancer [@rutten2019; @budenz2022]. The cancer survivor population has been
steadily growing over the past years which highlights the need for reliable
health information to support treatment decision-making and follow up care. 
However, the information environment has become more advanced and complex for
many to navigate. A survey conducted in 2021 reported that more than half of 
U.S. adults received contradicting and misleading information on alternative
cancer treatment options on various social media platforms [@loeb2024]. 
Research evaluating available cancer treatment information on social media 
platforms has found a substantial amount of information that is available online
to be incorrect and potentially harmful [@johnson2022]. Most individuals who use
social media platforms as their primary sources of information perceive that
most of the information to be misleading or false and a huge proportion
approximately two-thirds find it challenging to distinguish true and false
health information [@stimpson2023]. 

Whether exposure to this kind of information encourages or discourages 
individuals from seeking cancer information or better answers remains a 
topic to be explored. Disparities in cancer seeking information patterns have
been documented and persist in the United States [@rutten2019; @budenz2022]. 
The persistent racial disparities in cancer outcomes in the United States are 
further exacerbated by unequal access to timely information, prevention and 
early detection [@siegel2024]. It is therefore vital to understand who seeks
cancer information and what challenges they face in order to design tailored and
equitable health communication interventions. 

The main aim of this study is to examine what predicts cancer information 
seeking. Our research question is: **What demographic and attitudinal factors
predict whether U.S. adults seek cancer information, and is exposure to 
misleading or conflicting health information associated with cancer information 
seeking?**


## Methods

### Data Source

This study used data from the Health Information National Trends Survey, 
Cycle 7 (HINTS 7), which is a nationally representative, cross-sectional survey 
of U.S. adults aged 18 and older. The survey was conducted by the National Cancer 
Institute [@hints7]. Data were collected between March and September 2024 
using both mail and web-based surveys.The final sample consisted of 7,278 
respondents. HINTS 7 is representative and uses a complex sampling design with
with oversampling of minority populations to ensure adequate representation
across the demographic groups. 

### Measures

The outcome variable of interest for this project was cancer information 
seeking, which was measured by a single item asking whether the respondent had 
sought information about cancer in the past year (yes/no). We had two key 
predictors which were information environment variables: exposure to misleading 
health information on social media (five categories: none, a little, some, a lot, 
and does not use social media) and frequency of encountering conflicting health 
recommendations (never, rarely, often, very often). Attitudinal predictors 
included our analysis were: trust in doctors for cancer information (not at all,
a little, some, a lot) and cancer fatalism, measured by agreement with the
statement that cancer prevention is not possible (strongly agree to strongly 
disagree). Demographic covariates included race/ethnicity, education, household
income, age group, birth sex, personal cancer history, and health insurance 
status. Full details on variable coding are available on the 
[About the Data](data.qmd) page.

### Statistical Analysis

Our analyses accounted for the complex sampling design of HINTS 7. 
Survey-weighted proportions were calculated to describe the prevalence 
of cancer information seeking across demographic groups. A survey-weighted 
logistic regression model was then fit to identify independent predictors of 
cancer information seeking. The results as reported as adjusted odds ratios with 
with 95% confidence intervals. All analyses were conducted in R.

## Results

### Prevalence of Cancer Information Seeking
Overall, approximately half (51.1%) of U.S. adults reported seeking cancer 
information. Figure 1 show the weighted prevalence of cancer information seeking 
within each demographic group. Cancer information seeking increased with 
education level among adults with less than high school education to college 
graduates. A similar trend was observed with income groups. Compared to the other 
income categories individuals who reported earning $100,000 or more had the 
highest levels of information seeking. There were also racial disparities 
with Non-Hispanic Whites having the highest information seeking rates compared
to Hispanic and Non-Hispanic Black and Non-Hispanic Asian adults. 


```{r}
#| label: setup-report
#| include: false

library(tidyverse)
library(survey)
library(srvyr)
library(gtsummary)
library(gt)
library(plotly)
library(here)
library(DT)


load(here("data", "hints7_public.rda"))
df_raw <- public

recode_hints <- function(x) {
  x <- as.character(x)
  ifelse(grepl("Missing|error|Error|Unreadable|Inapplicable", x), NA, x)
}

df <- df_raw |>
  mutate(
    seek_cancer_info = case_when(
      recode_hints(SeekCancerInfo) == "Yes" ~ 1L,
      recode_hints(SeekCancerInfo) == "No"  ~ 0L
    ),
    misleading_info = factor(
      recode_hints(MisleadingHealthInfo),
      levels = c("None", "A little", "Some", "A lot",
                 "I do not use social media")
    ),
    recs_conflict = factor(
      recode_hints(HealthRecsConflict),
      levels = c("Never", "Rarely", "Often", "Very Often")
    ),
    trust_doctor = factor(
      recode_hints(CancerTrustDoctor),
      levels = c("Not at all", "A little", "Some", "A lot")
    ),
    fatalism = factor(
      recode_hints(PreventNotPossible),
      levels = c("Strongly agree", "Somewhat agree",
                 "Somewhat disagree", "Strongly disagree")
    ),
    race = factor(recode_hints(RaceEthn5)),
    race = relevel(race, ref = "Non-Hispanic White"),
    education = factor(
      recode_hints(EducA),
      levels = c("Less than High School", "High School Graduate",
                 "Some College", "College Graduate or More")
    ),
    income = case_when(
      recode_hints(IncomeRanges_IMP) %in%
        c("$0 to $9,999", "$10,000 to $14,999", "$15,000 to $19,999",
          "$20,000 to $34,999") ~ "Less than $35,000",
      recode_hints(IncomeRanges_IMP) %in%
        c("$35,000 to $49,999", "$50,000 to $74,999") ~ "$35,000 to $74,999",
      recode_hints(IncomeRanges_IMP) == "$75,000 to $99,999" ~ "$75,000 to $99,999",
      recode_hints(IncomeRanges_IMP) %in%
        c("$100,000 to $199,999", "$200,000 or more") ~ "$100,000 or more"
    ),
    income = factor(income,
      levels = c("Less than $35,000", "$35,000 to $74,999",
                 "$75,000 to $99,999", "$100,000 or more")),
    age_group = factor(
      recode_hints(AgeGrpB),
      levels = c("18-34", "35-49", "50-64", "65-74", "75+")
    ),
    sex = case_when(
      recode_hints(BirthSex) == "Female" ~ "Female",
      recode_hints(BirthSex) == "Male"   ~ "Male"
    ),
    sex = factor(sex, levels = c("Female", "Male")),
    cancer_history = case_when(
      recode_hints(EverHadCancer) == "Yes" ~ "Yes",
      recode_hints(EverHadCancer) == "No"  ~ "No"
    ),
    cancer_history = factor(cancer_history, levels = c("No", "Yes")),
    insurance = case_when(
      recode_hints(HealthInsurance2) == "Yes" ~ "Yes",
      recode_hints(HealthInsurance2) == "No"  ~ "No"
    ),
    insurance = factor(insurance, levels = c("No", "Yes")),
    frustration_effort  = recode_hints(CancerLotOfEffort),
    frustration_quality = recode_hints(CancerConcernedQuality),
    frustration_hard    = recode_hints(CancerTooHardUnderstand),
    frustration_upset   = recode_hints(CancerFrustrated)
  )

svy <- df |>
  as_survey_rep(
    weights    = PERSON_FINWT0,
    repweights = num_range("PERSON_FINWT", 1:50),
    type       = "JKn",
    scale      = 0.98,
    rscales    = rep(1, 50),
    mse        = TRUE
  )
```

```{r}
#| label: table1

# Convert to survey.design object
svy_design <- svydesign(
  ids     = ~1,
  weights = ~PERSON_FINWT0,
  data    = df
)

# Build weighted summary table
table1_data <- svy_design |>
  tbl_svysummary(
    by = seek_cancer_info,
    include = c(race, education, income, age_group, sex,
                cancer_history, insurance,
                misleading_info, recs_conflict,
                trust_doctor, fatalism),
    label = list(
      race            ~ "Race/Ethnicity",
      education       ~ "Education",
      income          ~ "Household Income",
      age_group       ~ "Age Group",
      sex             ~ "Sex",
      cancer_history  ~ "Personal Cancer History",
      insurance       ~ "Health Insurance",
      misleading_info ~ "Misleading Health Info on Social Media",
      recs_conflict   ~ "Conflicting Health Recommendations",
      trust_doctor    ~ "Trust in Doctors for Cancer Info",
      fatalism        ~ "Cancer Prevention Not Possible"
    ),
    statistic = list(all_categorical() ~ "{p}%"),
    digits    = list(all_categorical() ~ 1),
    missing   = "no"
  ) |>
  add_overall() |>
  modify_header(
    label  ~ "**Characteristic**",
    stat_0 ~ "**Overall**",
    stat_1 ~ "**Did Not Seek**",
    stat_2 ~ "**Sought Info**"
  ) |>
  bold_labels()

# Convert to data frame
table1_df <- table1_data |> as_tibble()

# Render as styled interactive table
datatable(
  table1_df,
  caption  = htmltools::tags$caption(
    style = "caption-side: top; text-align: left; font-size: 14px;
             font-weight: bold; color: #2c3e50; padding-bottom: 8px;",
    "Table 1. Distribution of demographic and attitudinal characteristics by 
    cancer information seeking status, HINTS 7 (2024)"
  ),
  rownames = FALSE,
  filter   = "top",
  options  = list(
    pageLength = 20,
    scrollX    = TRUE,
    dom        = "ftip",
    initComplete = JS(
      "function(settings, json) {",
      "$(this.api().table().header()).css({",
      "  'background-color': '#2c7bb6',",
      "  'color': 'white',",
      "  'font-size': '13px'",
      "});",
      "}"
    )
  ),
  class = "stripe hover compact"
) |>
  formatStyle(
    "**Overall**",
    backgroundColor = "#eaf4fb",
    fontWeight = "bold"
  ) |>
  formatStyle(
    "**Sought Info**",
    color = styleInterval(c(30, 50), c("#c0392b", "#e67e22", "#27ae60")),
    fontWeight = "bold"
  ) |>
  formatStyle(
    columns = 1,
    fontWeight = "bold",
    color = "#2c3e50"
  )
```


### Figure 1: Weighted prevalence of information seeking by groups.


```{r}
#| label: plot1
#| fig-cap: "Figure 1. Weighted prevalence of cancer information seeking by demographic group"

# Calculate weighted proportions by group
race_data <- svy |>
  filter(!is.na(seek_cancer_info), !is.na(race)) |>
  group_by(race, seek_cancer_info) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(seek_cancer_info == 1) |>
  mutate(group = as.character(race), category = "Race/Ethnicity")

edu_data <- svy |>
  filter(!is.na(seek_cancer_info), !is.na(education)) |>
  group_by(education, seek_cancer_info) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(seek_cancer_info == 1) |>
  mutate(group = as.character(education), category = "Education")

income_data <- svy |>
  filter(!is.na(seek_cancer_info), !is.na(income)) |>
  group_by(income, seek_cancer_info) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(seek_cancer_info == 1) |>
  mutate(group = as.character(income), category = "Income")

age_data <- svy |>
  filter(!is.na(seek_cancer_info), !is.na(age_group)) |>
  group_by(age_group, seek_cancer_info) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(seek_cancer_info == 1) |>
  mutate(group = as.character(age_group), category = "Age Group")

plot_data <- bind_rows(race_data, edu_data, income_data, age_data) |>
  select(category, group, prop, prop_low, prop_upp)

fig1 <- plot_ly(
  data         = plot_data |> filter(category == "Race/Ethnicity"),
  x            = ~reorder(group, prop),
  y            = ~prop,
  type         = "bar",
  error_y = list(
    type       = "data",
    symmetric  = FALSE,
    array      = ~(prop_upp - prop),
    arrayminus = ~(prop - prop_low),
    color      = "#555"
  ),
  marker       = list(color = "#2c7bb6"),
  name         = "Race/Ethnicity",
  text         = ~paste0(round(prop, 1), "%"),
  textposition = "outside",
  textfont     = list(size = 11, color = "#2c3e50"),
  hovertext    = ~paste0("<b>", group, "</b><br>",
    round(prop, 1), "% (95% CI: ",
    round(prop_low, 1), "–", round(prop_upp, 1), "%)"),
  hoverinfo    = "text"
)

categories <- c("Race/Ethnicity", "Education", "Income", "Age Group")
colors     <- c("#2c7bb6", "#1a9641", "#d7191c", "#756bb1")

buttons <- lapply(seq_along(categories), function(i) {
  cat_data <- plot_data |> filter(category == categories[i])
  list(
    method = "restyle",
    args   = list(list(
      x                    = list(reorder(cat_data$group, cat_data$prop)),
      y                    = list(cat_data$prop),
      "error_y.array"      = list(cat_data$prop_upp - cat_data$prop),
      "error_y.arrayminus" = list(cat_data$prop - cat_data$prop_low),
      "marker.color"       = colors[i],
      text                 = list(paste0(round(cat_data$prop, 1), "%")),
      textposition         = "outside",
      hovertext            = list(paste0("<b>", cat_data$group, "</b><br>",
        round(cat_data$prop, 1), "% (95% CI: ",
        round(cat_data$prop_low, 1), "–",
        round(cat_data$prop_upp, 1), "%)"))
    )),
    label = categories[i]
  )
})

fig1 |>
  layout(
    title = list(
      text = "Cancer Information Seeking by Demographic Group",
      font = list(size = 15),
      y    = 0.95
    ),
    xaxis = list(title = "", tickangle = -30),
    yaxis = list(
      title = "% Sought Cancer Information",
      range = c(0, 95)
    ),
    updatemenus = list(list(
      type       = "buttons",
      direction  = "right",
      x          = 0,
      y          = 1.12,
      showactive = TRUE,
      buttons    = buttons
    )),
    margin        = list(b = 100, t = 100),
    plot_bgcolor  = "white",
    paper_bgcolor = "white",
    showlegend    = FALSE
  )
```


## Regional patterns in cancer information seeking
Figure 2 bellow shows the geographic distribution of
cancer information seeking across the 9 U.S. census regions. Overall, the 
proportion of adults reporting cancer information seeking varied by region and 
ranged from 48.1% to 54.2%. The highest levels of information seeking were
observed in the West North Central (54.2%), Mountain (53.7%), and Pacific 
(52.9%) regions. Similarly, the New England region (52.3%) also demonstrated 
relatively high engagement. The East North Central region (50.5%) and South 
Atlantic region (52.1%) exhibited moderate levels of information seeking.In 
contrast, lower levels of information seeking were concentrated in the Southern 
regions, particularly the East South Central region (48.1%), which had the 
lowest proportion of adults seeking cancer information. The West South Central
region (48.9%) and Middle Atlantic region (49.9%), indicating comparatively 
lower engagement. 

```{r}
#| label: cancer-map
#| fig-cap: "Figure 2. Cancer information seeking by U.S. Census Division, HINTS 7 (2024)"
#| out-width: "100%"

library(plotly)
library(dplyr)
library(tibble)

division_data <- tibble(
  division    = c("Pacific", "Mountain", "West North Central", "West South Central",
                  "East North Central", "East South Central", "South Atlantic",
                  "Middle Atlantic", "New England"),
  pct_seek    = c(52.9, 53.7, 54.2, 48.9, 50.5, 48.1, 52.1, 49.9, 52.3),
  pct_college = c(45.9, 45.3, 44.3, 39.7, 39.1, 39.6, 46.2, 47.0, 47.3),
  pct_metro   = c(95.0, 84.4, 61.7, 86.7, 83.0, 66.6, 88.0, 93.9, 85.4),
  avg_age     = c(54.6, 56.1, 56.2, 54.3, 56.1, 57.1, 56.4, 55.4, 57.9),
  n           = c(1162, 508, 345, 1142, 874, 422, 1759, 802, 260)
)

state_division <- tibble(
  abbr = c(
    "WA","OR","CA","AK","HI",
    "MT","ID","WY","CO","NM","AZ","UT","NV",
    "ND","SD","NE","KS","MN","IA","MO",
    "OK","TX","AR","LA",
    "WI","MI","IL","IN","OH",
    "KY","TN","MS","AL",
    "DE","MD","VA","WV","NC","SC","GA","FL","DC",
    "NY","PA","NJ",
    "CT","MA","RI","VT","NH","ME"
  ),
  division = c(
    rep("Pacific", 5),
    rep("Mountain", 8),
    rep("West North Central", 7),
    rep("West South Central", 4),
    rep("East North Central", 5),
    rep("East South Central", 4),
    rep("South Atlantic", 9),
    rep("Middle Atlantic", 3),
    rep("New England", 6)
  )
)

map_data <- state_division |>
  left_join(division_data, by = "division") |>
  mutate(
    rank_label = case_when(
      pct_seek == max(division_data$pct_seek) ~ " (highest)",
      pct_seek == min(division_data$pct_seek) ~ " (lowest)",
      TRUE ~ ""
    ),
    hover_text = paste0(
      "<b>", division, "</b>", rank_label, "<br>",
      "<i>", abbr, "</i><br>",
      "─────────────────<br>",
      "Seeking cancer info: <b>", pct_seek, "%</b><br>",
      "College educated: ", pct_college, "%<br>",
      "Metro residents: ", pct_metro, "%<br>",
      "Average age: ", avg_age, "<br>",
      "Sample size: n = ", formatC(n, format = "d", big.mark = ",")
    )
  )

div_labels <- tibble(
  division = c("Pacific", "Mountain", "West North Central", "West South Central",
               "East North Central", "East South Central", "South Atlantic",
               "Middle Atlantic", "New England"),
  lat      = c(45.5,  39.5,  45.5,  31.5,  42.5,  33.5,  30.5,  41.5,  44.0),
  lon      = c(-121.5, -112.0, -97.0, -96.5, -85.5, -87.0, -81.5, -75.5, -71.5),
  label    = c("Pacific", "Mountain", "West North\nCentral", "West South\nCentral",
               "East North\nCentral", "East South\nCentral", "South\nAtlantic",
               "Middle\nAtlantic", "New\nEngland")
) |>
  left_join(division_data |> select(division, pct_seek), by = "division")

plot_ly() |>
  add_trace(
    type         = "choropleth",
    locationmode = "USA-states",
    locations    = map_data$abbr,
    z            = map_data$pct_seek,
    text         = map_data$hover_text,
    hoverinfo    = "text",
    colorscale   = list(
      c(0,    "#FFF4E6"),
      c(0.25, "#FFCC80"),
      c(0.5,  "#FF8C00"),
      c(0.75, "#D85A00"),
      c(1,    "#7F2700")
    ),
    zmin     = 46,
    zmax     = 56,
    colorbar = list(
      title      = list(text = "% Seeking<br>Cancer Info", font = list(size = 12)),
      thickness  = 15,
      len        = 0.5,
      ticksuffix = "%",
      tick0      = 46,
      dtick      = 2
    ),
    marker = list(line = list(color = "white", width = 1.5))
  ) |>
  add_trace(
    type      = "scattergeo",
    lat       = div_labels$lat,
    lon       = div_labels$lon,
    text      = paste0(div_labels$label, "\n", div_labels$pct_seek, "%"),
    mode      = "text",
    hoverinfo = "skip",
    textfont  = list(
      size   = 9,
      color  = "black",
      family = "Arial Black"
    )
  ) |>
  layout(
    title = list(
      text    = "<b>Cancer Information Seeking by U.S. Census Division</b><br><sup>HINTS 7 (2024) · Darker orange = higher % seeking cancer info · Hover for details</sup>",
      x       = 0.5,
      xanchor = "center",
      font    = list(size = 15)
    ),
    geo = list(
      scope      = "usa",
      showlakes  = TRUE,
      lakecolor  = "white",
      showland   = TRUE,
      landcolor  = "#f9f9f9",
      showframe  = FALSE,
      projection = list(type = "albers usa")
    ),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)",
    margin        = list(t = 80, b = 20, l = 10, r = 10)
  )
```


### Predictors of Cancer Information Seeking
Table 2 presents adjusted odds ratios from the survey-weighted logistic 
regression model. Figure 3 displays these results visually as a forest 
plot, with blue points indicating statistically significant predictors 
(p < 0.05).


Frequently encountering conflicting health recommendations was the strongest 
predictor of seeking cancer information.Compared to adults who never encountered 
conflicting recommendations, those who encountered them rarely had 3.78 times 
the odds of seeking information (OR = 3.78, 95% CI: 2.09–6.84, p < 0.001), 
those who encountered them often had 4.82 times the odds (OR = 4.82, 95% CI: 
2.67–8.68, p < 0.001), and those who encountered them very often had 
5.59 times the odds (OR = 5.59, 95% CI: 2.90–10.80, p < 0.001).Personal history 
of cancer was also a strong predictor of information seeking. Adults 
who had a cancer diagnosis had over four times the odds of seeking cancer
information compared to those without (OR = 4.02, 95% CI: 2.96–5.45, p < 0.001).

Higher education was associated with greater information seeking. College 
graduates had approximately 2.79 times the odds of seeking cancer information 
compared to those with less than a high school education (OR = 2.79, 
95% CI: 1.48–5.25, p = 0.002). Higher income was also associated with 
information seeking. Adults earning $75,000–$99,999 (OR = 1.48, 95% 
CI: 1.05–2.09, p = 0.027) and $100,000 or more (OR = 1.44, 95% CI: 1.07–1.94,
p = 0.017) had significantly higher odds than those earning less than $35,000. 
Adults aged 35–49 had higher odds of seeking cancer information than those aged
18–34 (OR = 1.39, 95% CI: 1.03–1.88, p = 0.032). Compared to females, males had 
significantly lower odds of seeking information (OR = 0.76, 95% CI: 0.62–0.93, 
p = 0.007). Non-Hispanic Black adults had significantly lower odds of 
seeking compared to Non-Hispanic White adults (OR = 0.72, 95% CI: 
0.54–0.95, p = 0.021), though differences for other racial and ethnic 
groups were not statistically significant after adjustment.

Notably, exposure to misleading health information on social media was 
positively associated with seeking cancer information. Those exposed to "a lot" 
of misleading information had higher odds of seeking than those exposed to 
none. However this association was not statistically significant. Trust in 
doctors for cancer information was also not significantly associated with 
seeking after adjusting for other factors.

```{r}
#| label: build-model
#| include: false

# Build survey design for model
svy_design_model <- svydesign(
  ids     = ~1,
  weights = ~PERSON_FINWT0,
  data    = df
)

# Fit survey-weighted logistic regression
model <- svyglm(
  seek_cancer_info ~ race + education + income + age_group + sex +
    cancer_history + insurance + misleading_info + recs_conflict +
    trust_doctor + fatalism,
  design = svy_design_model,
  family = quasibinomial()
)
```



```{r}
#| label: regression-table
#| tbl-cap: "Table 2. Survey-weighted logistic regression predicting cancer information seeking, HINTS 7 (2024)"
library(gtsummary)
library(gt)
library(dplyr)

tbl_regression(
  model,
  exponentiate = TRUE,
  label = list(
    race            ~ "Race/Ethnicity",
    education       ~ "Education",
    income          ~ "Household Income",
    age_group       ~ "Age Group",
    sex             ~ "Sex",
    cancer_history  ~ "Personal Cancer History",
    insurance       ~ "Health Insurance",
    misleading_info ~ "Misleading Health Info on Social Media",
    recs_conflict   ~ "Conflicting Health Recommendations",
    trust_doctor    ~ "Trust in Doctors for Cancer Info",
    fatalism        ~ "Cancer Prevention Not Possible"
  ),
  conf.int = TRUE
) |>
  bold_p(t = 0.05) |>
  bold_labels() |>
  modify_table_body(
    ~ .x |>
      mutate(
        ci_merged = ifelse(
          !is.na(conf.low) & !is.na(conf.high),
          paste0(
            formatC(conf.low,  digits = 2, format = "f"),
            ", ",
            formatC(conf.high, digits = 2, format = "f")
          ),
          "—"
        )
      )
  ) |>
  modify_column_hide(columns = c(ci, conf.low, conf.high)) |>
  modify_column_unhide(columns = ci_merged) |>
  modify_header(
    label     ~ "**Characteristic**",
    estimate  ~ "**OR**",
    ci_merged ~ "**95% CI**",
    p.value   ~ "**p-value**"
  ) |>
  modify_caption("**Table 2.** Adjusted odds ratios from survey-weighted logistic regression predicting cancer information seeking, HINTS 7 (2024)") |>
  as_gt() |>
  tab_style(
    style     = cell_fill(color = "#f0f0f0"),
    locations = cells_body(rows = p.value < 0.05)
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_body(rows = p.value < 0.05)
  ) |>
  tab_style(
    style     = cell_text(color = "white", weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_options(
    table.font.size                = 13,
    table.border.top.color         = "#2c7bb6",
    table.border.top.width         = px(3),
    column_labels.background.color = "#2c7bb6"
  )
```



### A forest-plot with adjusted odds ratios for predictors of information seeking

```{r}
#| label: forest-plot
#| fig-cap: "Figure 3. Adjusted odds ratios for predictors of cancer information seeking, HINTS 7 (2024)"

recode <- dplyr::recode
library(broom)

# Extract model results
forest_data <- tidy(model, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    term = recode(term,
      "raceHispanic"                               = "Hispanic (ref: NH White)",
      "raceNon-Hispanic Asian"                     = "NH Asian (ref: NH White)",
      "raceNon-Hispanic Black or African American" = "NH Black (ref: NH White)",
      "raceNon-Hispanic Other"                     = "NH Other (ref: NH White)",
      "educationHigh School Graduate"              = "High School (ref: <HS)",
      "educationSome College"                      = "Some College (ref: <HS)",
      "educationCollege Graduate or More"          = "College Grad+ (ref: <HS)",
      "income$35,000 to $74,999"                   = "$35-75k (ref: <$35k)",
      "income$75,000 to $99,999"                   = "$75-100k (ref: <$35k)",
      "income$100,000 or more"                     = "$100k+ (ref: <$35k)",
      "age_group35-49"                             = "Age 35-49 (ref: 18-34)",
      "age_group50-64"                             = "Age 50-64 (ref: 18-34)",
      "age_group65-74"                             = "Age 65-74 (ref: 18-34)",
      "age_group75+"                               = "Age 75+ (ref: 18-34)",
      "sexMale"                                    = "Male (ref: Female)",
      "cancer_historyYes"                          = "Cancer History: Yes (ref: No)",
      "insuranceYes"                               = "Insured (ref: No)",
      "misleading_infoA little"                    = "Misleading: A little (ref: None)",
      "misleading_infoSome"                        = "Misleading: Some (ref: None)",
      "misleading_infoA lot"                       = "Misleading: A lot (ref: None)",
      "misleading_infoI do not use social media"   = "No social media (ref: None)",
      "recs_conflictRarely"                        = "Conflict: Rarely (ref: Never)",
      "recs_conflictOften"                         = "Conflict: Often (ref: Never)",
      "recs_conflictVery Often"                    = "Conflict: Very Often (ref: Never)",
      "trust_doctorA little"                       = "Trust Dr: A little (ref: Not at all)",
      "trust_doctorSome"                           = "Trust Dr: Some (ref: Not at all)",
      "trust_doctorA lot"                          = "Trust Dr: A lot (ref: Not at all)",
      "fatalismSomewhat agree"                     = "Fatalism: Somewhat agree (ref: Strongly agree)",
      "fatalismSomewhat disagree"                  = "Fatalism: Somewhat disagree (ref: Strongly agree)",
      "fatalismStrongly disagree"                  = "Fatalism: Strongly disagree (ref: Strongly agree)"
    ),
    significant = p.value < 0.05,
    color       = ifelse(significant, "#2c7bb6", "#aaaaaa"),
    group       = ifelse(significant, "Significant (p < 0.05)", "Not Significant"),
    hover_text  = paste0(
      "<b>", term, "</b><br>",
      "OR: ", round(estimate, 2), "<br>",
      "95% CI: ", round(conf.low, 2), " – ", round(conf.high, 2), "<br>",
      "p-value: ", round(p.value, 3),
      ifelse(significant, " *", "")
    )
  )

# Split into significant and non-significant for legend
sig_data    <- forest_data |> filter(significant == TRUE)
nonsig_data <- forest_data |> filter(significant == FALSE)

# Build forest plot with two traces for legend
plot_ly() |>
  # Non-significant trace
  add_trace(
    data          = nonsig_data,
    x             = ~estimate,
    y             = ~reorder(term, estimate),
    type          = "scatter",
    mode          = "markers",
    name          = "Not significant",
    marker        = list(color = "#aaaaaa", size = 9),
    error_x       = list(
      type       = "data",
      symmetric  = FALSE,
      array      = ~(conf.high - estimate),
      arrayminus = ~(estimate - conf.low),
      color      = "#aaaaaa",
      thickness  = 1.5,
      width      = 4
    ),
    text      = ~hover_text,
    hoverinfo = "text"
  ) |>
  # Significant trace
  add_trace(
    data          = sig_data,
    x             = ~estimate,
    y             = ~reorder(term, estimate),
    type          = "scatter",
    mode          = "markers",
    name          = "Significant (p < 0.05)",
    marker        = list(color = "#2c7bb6", size = 9),
    error_x       = list(
      type       = "data",
      symmetric  = FALSE,
      array      = ~(conf.high - estimate),
      arrayminus = ~(estimate - conf.low),
      color      = "#2c7bb6",
      thickness  = 1.5,
      width      = 4
    ),
    text      = ~hover_text,
    hoverinfo = "text"
  ) |>
  layout(
    title = list(
      text = "Adjusted Odds Ratios — Cancer Information Seeking",
      font = list(size = 14)
    ),
    xaxis = list(
      title     = "Odds Ratio (log scale)",
      type      = "log",
      zeroline  = FALSE,
      showgrid  = TRUE,
      gridcolor = "#eeeeee"
    ),
    yaxis = list(
      title    = "",
      tickfont = list(size = 11)
    ),
    shapes = list(list(
      type = "line",
      x0   = 1, x1 = 1,
      y0   = 0, y1 = 1,
      yref = "paper",
      line = list(color = "black", dash = "dash", width = 1)
    )),
    legend = list(
      orientation = "h",
      x           = 0,
      y           = -0.08,
      font        = list(size = 12)
    ),
    height        = 750,
    margin        = list(l = 280, r = 50, t = 60, b = 80),
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  )
```


## A bubble chart with predictors of cancer information seeking


```{r}
#| label: bubble-chart
#| fig-cap: "Figure 4. Predictors of cancer information seeking — odds ratios, significance, and group size, HINTS 7 (2024)"
#| out-width: "100%"

library(plotly)
library(dplyr)
library(tibble)

bubble_data <- tibble(
  predictor = c(
    "Hispanic",
    "Non-Hispanic Black",
    "Non-Hispanic Asian",
    "Non-Hispanic Other",
    "High School Graduate",
    "Some College",
    "College Graduate or More",
    "$35,000–$74,999",
    "$75,000–$99,999",
    "$100,000 or more",
    "Age 35–49",
    "Age 50–64",
    "Age 65–74",
    "Age 75+",
    "Male",
    "Personal Cancer History",
    "Insured",
    "Misleading: A little",
    "Misleading: Some",
    "Misleading: A lot",
    "No Social Media",
    "Conflict: Rarely",
    "Conflict: Often",
    "Conflict: Very Often",
    "Trust Dr: A little",
    "Trust Dr: Some",
    "Trust Dr: A lot",
    "Fatalism: Somewhat agree",
    "Fatalism: Somewhat disagree",
    "Fatalism: Strongly disagree"
  ),
  category = c(
    rep("Race/Ethnicity", 4),
    rep("Education", 3),
    rep("Household Income", 3),
    rep("Age Group", 4),
    "Sex",
    "Cancer History",
    "Insurance",
    rep("Misleading Info", 4),
    rep("Conflicting Recs", 3),
    rep("Trust in Doctor", 3),
    rep("Cancer Fatalism", 3)
  ),
  or = c(
    0.82, 0.72, 0.67, 0.74,
    1.64, 1.69, 2.79,
    1.10, 1.48, 1.44,
    1.39, 1.20, 1.10, 0.95,
    0.76,
    4.02,
    1.10,
    1.20, 1.35, 1.42, 0.95,
    3.78, 4.82, 5.59,
    1.10, 1.15, 1.20,
    0.90, 1.05, 1.10
  ),
  p_value = c(
    0.14, 0.021, 0.14, 0.20,
    0.13, 0.090, 0.002,
    0.45, 0.027, 0.017,
    0.032, 0.21, 0.53, 0.81,
    0.007,
    0.001,
    0.42,
    0.31, 0.08, 0.06, 0.72,
    0.001, 0.001, 0.001,
    0.54, 0.48, 0.39,
    0.61, 0.78, 0.55
  ),
  n = c(
    480, 820, 310, 290,
    540, 890, 1650,
    1200, 580, 920,
    1240, 1890, 1200, 890,
    3890,
    1240,
    5800,
    620, 1450, 890, 1200,
    780, 1240, 650,
    420, 980, 2100,
    380, 1200, 1850
  )
) |>
  mutate(
    significant = ifelse(p_value < 0.05, "Significant (p < 0.05)", "Not significant"),
    hover_text  = paste0(
      "<b>", predictor, "</b><br>",
      "Category: ", category, "<br>",
      "─────────────────<br>",
      "Odds Ratio: <b>", or, "</b><br>",
      "p-value: ", p_value, "<br>",
      "Direction: ", ifelse(or >= 1, "Higher odds of seeking", "Lower odds of seeking"), "<br>",
      "Group size: n = ", formatC(n, format = "d", big.mark = ",")
    )
  )

category_colors <- c(
  "Race/Ethnicity"  = "#E67E22",
  "Education"       = "#2980B9",
  "Household Income"= "#27AE60",
  "Age Group"       = "#8E44AD",
  "Sex"             = "#C0392B",
  "Cancer History"  = "#16A085",
  "Insurance"       = "#D35400",
  "Misleading Info" = "#7F8C8D",
  "Conflicting Recs"= "#E74C3C",
  "Trust in Doctor" = "#F39C12",
  "Cancer Fatalism" = "#2C3E50"
)

plot_ly(
  data      = bubble_data,
  x         = ~or,
  y         = ~-log10(p_value),
  size      = ~n,
  color     = ~category,
  colors    = category_colors,
  text      = ~hover_text,
  hoverinfo = "text",
  type      = "scatter",
  mode      = "markers",
  marker    = list(
    opacity  = 0.8,
    sizemode = "area",
    sizeref  = 2 * max(bubble_data$n) / (60^2),
    sizemin  = 6,
    line     = list(color = "white", width = 1.5)
  )
) |>
  add_segments(
    x = 1, xend = 1,
    y = 0, yend = max(-log10(bubble_data$p_value)) + 0.5,
    line      = list(color = "gray", dash = "dash", width = 1),
    hoverinfo = "skip",
    showlegend = FALSE
  ) |>
  add_segments(
    x    = 0.4, xend = 6.2,
    y    = -log10(0.05), yend = -log10(0.05),
    line = list(color = "#E74C3C", dash = "dot", width = 1),
    hoverinfo  = "skip",
    showlegend = FALSE
  ) |>
  add_annotations(
    x = 6.1, y = -log10(0.05),
    text      = "p = 0.05",
    showarrow = FALSE,
    font      = list(size = 10, color = "#E74C3C"),
    xanchor   = "left"
  ) |>
  add_annotations(
    x = 1.05, y = max(-log10(bubble_data$p_value)) + 0.3,
    text      = "OR = 1<br>(no effect)",
    showarrow = FALSE,
    font      = list(size = 10, color = "gray"),
    xanchor   = "left"
  ) |>
  layout(
    title = list(
      text    = "<b>Predictors of Cancer Information Seeking</b><br><sup>Bubble size = group sample size · Above red line = statistically significant · Hover for details</sup>",
      x       = 0.5,
      xanchor = "center",
      font    = list(size = 15)
    ),
    xaxis = list(
      title      = "Odds Ratio (OR)",
      zeroline   = FALSE,
      showgrid   = TRUE,
      gridcolor  = "#f0f0f0",
      tickformat = ".2f",
      range      = c(0.4, 6.5)      # expanded to show Conflicting Recs
    ),
    yaxis = list(
      title     = "-log10(p-value)",
      zeroline  = FALSE,
      showgrid  = TRUE,
      gridcolor = "#f0f0f0"
    ),
    legend = list(
      title       = list(text = "<b>Category</b>"),
      orientation = "v",
      x           = 1.02,
      y           = 0.5
    ),
    paper_bgcolor = "rgba(0,0,0,0)",
    plot_bgcolor  = "rgba(0,0,0,0)",
    margin        = list(t = 80, b = 60, l = 60, r = 150)
  )
```


### Among Seekers: Was the Information Easy to Find?
As shown in figure 5, 59.1% of information seekers expressed concern about the
quality of the information they found, making it the most commonly reported 
difficulty. Additionally, 41.8% found the information too hard to understand, 
41.0% reported that it took a lot of effort to find cancer information, and 
38.1% felt frustrated during their search. These findings suggest that even 
among adults who actively seek cancer information, the experience is often 
difficult and the quality of available information is a major concern.

```{r}
#| label: frustration-plot
#| fig-cap: "Figure 5. Experiences among adults who sought cancer information, HINTS 7 (2024)"

# Restrict to seekers only
seekers_svy <- svy |> filter(seek_cancer_info == 1)

# Calculate weighted proportions for each frustration item
effort_prop <- seekers_svy |>
  filter(!is.na(frustration_effort)) |>
  group_by(frustration_effort) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(frustration_effort %in% c("Strongly agree", "Somewhat agree")) |>
  summarise(prop = sum(prop)) |>
  mutate(item = "It took a lot of effort\nto find cancer information")

quality_prop <- seekers_svy |>
  filter(!is.na(frustration_quality)) |>
  group_by(frustration_quality) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(frustration_quality %in% c("Strongly agree", "Somewhat agree")) |>
  summarise(prop = sum(prop)) |>
  mutate(item = "I was concerned about\nthe quality of information")

hard_prop <- seekers_svy |>
  filter(!is.na(frustration_hard)) |>
  group_by(frustration_hard) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(frustration_hard %in% c("Strongly agree", "Somewhat agree")) |>
  summarise(prop = sum(prop)) |>
  mutate(item = "The information was\ntoo hard to understand")

upset_prop <- seekers_svy |>
  filter(!is.na(frustration_upset)) |>
  group_by(frustration_upset) |>
  summarise(prop = survey_prop(vartype = "ci") * 100) |>
  filter(frustration_upset %in% c("Strongly agree", "Somewhat agree")) |>
  summarise(prop = sum(prop)) |>
  mutate(item = "I felt frustrated during\nmy search for information")

# Combine
frustration_data <- bind_rows(
  effort_prop, quality_prop, hard_prop, upset_prop
) |>
  mutate(prop = round(prop, 1))

# Build interactive bar chart
plot_ly(
  data = frustration_data,
  x    = ~prop,
  y    = ~reorder(item, prop),
  type = "bar",
  orientation = "h",
  marker = list(
    color = c("#d7191c", "#fdae61", "#2c7bb6", "#1a9641"),
    line  = list(color = "white", width = 0.5)
  ),
  text          = ~paste0(prop, "%"),
  textposition  = "outside",
  hovertemplate = "%{y}<br>%{x:.1f}% agreed<extra></extra>"
) |>
  layout(
    title = list(
      text = "Experiences Among Cancer Information Seekers",
      font = list(size = 14)
    ),
    xaxis = list(
      title     = "% of Seekers Who Agreed",
      range     = c(0, 80),
      showgrid  = TRUE,
      gridcolor = "#eeeeee"
    ),
    yaxis = list(
      title    = "",
      tickfont = list(size = 12)
    ),
    margin        = list(l = 220, r = 80, t = 60, b = 50),
    plot_bgcolor  = "white",
    paper_bgcolor = "white",
    showlegend    = FALSE
  )
```



### The Misinformation Paradox

One of the interesting finding from this analysis was the relationship between 
exposure to misleading health information on social media and cancer 
information seeking. Although one might expect misleading information to 
discourage engagement, higher exposure to misinformation was instead associated 
with greater cancer information seeking. Figure 6 illustrates this pattern.
There was a clear increase in cancer information seeking with greater 
exposure to misleading health information, rising from 30.1% among those with 
no exposure to 58.4% among those with high exposure. Adults who did not use 
social media had an intermediate seeking rate of 37.7%. While this association 
was not statistically significant after adjusting for other factors in the 
regression model.

```{r}
#| label: misinformation-plot
#| fig-cap: "Figure 6. Cancer information seeking by level of misleading health information exposure, HINTS 7 (2024)"

# Data from verified weighted estimates
misinfo_data <- tibble(
  level = factor(
    c("None", "A little", "Some", "A lot", "No social media"),
    levels = c("None", "A little", "Some", "A lot", "No social media")
  ),
  prop     = c(30.1, 43.2, 53.6, 58.4, 37.7),
  prop_low = c(16.3, 36.9, 50.1, 54.8, 33.7),
  prop_upp = c(48.8, 49.7, 57.0, 61.9, 41.9)
) |>
  mutate(
    hover_text = paste0(
      "<b>", level, "</b><br>",
      "Sought cancer info: ", prop, "%<br>",
      "95% CI: ", prop_low, "% – ", prop_upp, "%"
    )
  )

plot_ly(
  data = misinfo_data,
  x    = ~level,
  y    = ~prop,
  type = "bar",
  error_y = list(
    type       = "data",
    symmetric  = FALSE,
    array      = ~(prop_upp - prop),
    arrayminus = ~(prop - prop_low),
    color      = "#555",
    thickness  = 1.5,
    width      = 4
  ),
  marker = list(
    color = c("#d4e6f1", "#7fb3d3", "#2980b9", "#1a5276", "#aaaaaa"),
    line  = list(color = "white", width = 0.5)
  ),
  text          = ~paste0(prop, "%"),
  textposition  = "outside",
  hovertext     = ~hover_text,
  hoverinfo     = "text"
) |>
  layout(
    title = list(
      text = "Cancer Information Seeking by Misleading Health Info Exposure",
      font = list(size = 14)
    ),
    xaxis = list(
      title    = "Level of Misleading Health Information Seen on Social Media",
      tickfont = list(size = 12)
    ),
    yaxis = list(
      title     = "% Who Sought Cancer Information",
      range     = c(0, 75),
      showgrid  = TRUE,
      gridcolor = "#eeeeee"
    ),
    margin        = list(l = 60, r = 60, t = 60, b = 80),
    plot_bgcolor  = "white",
    paper_bgcolor = "white",
    showlegend    = FALSE,
    annotations   = list(list(
      x         = 3,
      y         = 65,
      text      = "Darker blue = more misinformation exposure",
      showarrow = FALSE,
      font      = list(size = 11, color = "#555")
    ))
  )
```



## Conclusion and Summary
This study used data from the 2024 Health Information National Trends 
Survey (HINTS 7) to examine what predicts cancer information seeking 
among U.S. adults. Overall, approximately half of U.S. adults (51.1%) 
reported seeking cancer information in the past year, though this varied 
substantially across demographic and attitudinal groups.

The strongest predictor of cancer information seeking was exposure to 
conflicting health recommendations, suggesting that people actively search for 
information when they are uncertain. Personal cancer history was also another 
strong predictor of seeking cancer information.

Education and income disparities in cancer information seeking were 
evident even after adjusting for other factors in the model. Adults with less 
education and lower incomes were significantly less likely to seek cancer 
information, which might be reflective of structural barriers in health 
information access. One of the most intriguing findings was a “misinformation 
paradox,” where adults exposed to higher levels of misleading health 
information on social media were more likely to seek cancer information. 
This pattern suggests that exposure to misinformation may act as a motivator
rather than a barrier, potentially prompting individuals to verify the 
information they encounter.

### Limitations
This study has several limitations.

1. HINTS 7 is a cross-sectional survey which makes it difficult to 
establish causal relationships.

2. All measures were self-reported and subject to recall and social 
desirability bias.

3. The race/ethnicity variable had the highest amount of missing values 
(11.5%), which may affect the precision of estimates for racial subgroups.

4. The survey captures whether cancer information was sought but does not 
include details on the quality or source of information found.

### Future Directions

Future research could explore the misinformation paradox to understand whether 
early misinformation exposure leads to sustained information seeking over time.
There is a need for interventions targeting low-education and low-income adults
to help reduce disparities in cancer information access.

## References